---
title: "LASSO_Final Project"
author: "Michelle_Abraham"
format: html 
echo: false
warning: false
message: false
editor: visual
---

```{r}
library(tidyverse)
library(janitor)
library(gt)
library(patchwork)
library(scales)
library(ggrepel)
library(glmnet)
```

```{r}
data <- read_csv("C:\\Users\\Michelle Abraham\\Downloads\\up_group2_data.csv")
data |> 
  mutate(
    student_teacher_ratio = (total_primary_school_students / total_primary_school_teachers),
    
    health_centre = ifelse(phc | sub_centre | chc, 1, 0),
    share_insurance_receiving = (gp_total_no_of_beneficiaries_rec / gp_total_no_of_eligible_benefici) * 100,
    share_icds_pregnant = (total_no_of_pregnant_women_recei / total_no_of_pregnant_women) * 100,
    share_below_6_immunized = (total_no_of_children_0_to_6_year / (total_female_child_age_bw_0_6 + total_male_child_age_bw_0_6)) * 100,
    share_underweight_newborn = (total_no_of_newly_born_underweig / total_no_of_newly_born_children) * 100,
    
    is_primary_school_with_toilet = ifelse(any_primary_sch_toilet >= 1, 1, 0),
    share_grd_pg = (total_grd_and_pg_in_village / total_population) * 100,
    
    share_hhd_piped = (total_hhd_having_piped_water_con / total_hhd) * 100,
    
    share_farming_activity = (total_hhd_engaged_in_farm_activi / total_hhd) * 100,
    share_hhd_ssh = (total_hhd_mobilized_into_shg / total_hhd) * 100,
    labour_budget_per_capita = (total_approved_labour_budget_for / total_population)
  ) |> 
  mutate(
    ln_labour_budget_per_capita = log(labour_budget_per_capita)
  ) -> data

data |> 
  mutate(across(everything(), ~ifelse(is.nan(.), NA, .))) |> 
  mutate(across(everything(), ~ifelse(is.infinite(.), NA, .))) -> data
variable_list = list(
  
  health = c("health_centre", "share_insurance_receiving", "share_icds_pregnant", "share_below_6_immunized", "share_underweight_newborn"),
  
  education = c("is_primary_school_with_electrici", "is_primary_school_have_drinking_", "is_primary_school_with_toilet", "availability_of_middle_school", "availability_of_high_school", "availability_of_govt_degree_coll", "share_grd_pg"),
  
  facilities = c("is_bank_available", "no_electricity", "share_hhd_piped", "is_village_connected_to_all_weat", "is_broadband_available"),
  
  other = c("share_farming_activity", "share_hhd_ssh", "public_info_board", "ln_labour_budget_per_capita")
)

variable_vec <- 
  variable_list |> 
  unlist() |> 
  unname()

tibble(variable_vec) |> 
  rename(variable = variable_vec) |> 
  mutate(
    long_name = case_when(
      variable == "health_centre" ~ "Health Centre (D)",
      variable == "share_insurance_receiving" ~ "Beneficiaries receiving PMJAY/State Health Insurance (%)",
      variable == "share_icds_pregnant" ~ "Pregnant Women receiving care under ICDS (%)",
      variable == "share_below_6_immunized" ~ "Children below 6 years Immunized (%)",
      variable == "share_underweight_newborn" ~ "Underweight Newborns (%)",
      
      variable == "is_primary_school_with_electrici" ~ "Electricity in Primary School (D)",
      variable == "is_primary_school_have_drinking_" ~ "Drinking Water in Primary School (D)",
      variable == "is_primary_school_with_toilet" ~ "Toilet in Primary School (D)",
      variable == "availability_of_middle_school" ~ "Middle School (D)",
      variable == "availability_of_high_school" ~ "High School (D)",
      variable == "availability_of_govt_degree_coll" ~ "Govt. Degree granting College (D)",
      variable == "share_grd_pg" ~ "Undergraduate & Postgraduate Population (%)",
      
      variable == "is_bank_available" ~ "Bank (D)",
      variable == "no_electricity" ~ "No Electricity (D)",
      variable == "share_hhd_piped" ~ "Share of Households with Piped Water (%)",
      variable == "is_village_connected_to_all_weat" ~ "Connection to All Weather Road (D)",
      variable == "is_broadband_available" ~ "Broadband/Internet Facilty (D)",
      
      variable == "share_farming_activity" ~ "Population in Farming (%)",
      variable == "share_hhd_ssh" ~ "Households mobilized into SHGs (%)",
      variable == "public_info_board" ~ "Public Information Board (D)",
      variable == "labour_budget_per_capita" ~ "Labour Budget per Capita",
      variable == "ln_labour_budget_per_capita" ~ "ln(Labour Budget per Capita)"
    )
  ) -> variable_long_names
give_long_name <- function(short_name){
  variable_long_names |> 
    filter(variable == short_name) |> 
    pull(long_name)
}

data |> 
  filter(student_teacher_ratio <= 500) -> data

set.seed(2025)

data |> 
  slice(0) -> test_data

data |> 
  distinct(district_name) |> 
  pull(district_name) -> district_vec

lapply(district_vec, function(x){
  district <- x
  
  data |> 
    filter(district_name == district) %>%
    slice_sample(n = round(0.3 * nrow(.)))
}) |> 
  bind_rows() -> test_data

test_data |> 
  pull(shrid2) -> test_data_shrid2

data |> 
  filter(!(shrid2 %in% test_data_shrid2)) -> training_data
district_wise_regression <- function(data, response_var, independent_var_vec){
  
  district_vec <- 
    data |> 
    distinct(district_name) |> 
    arrange(district_name) |> 
    pull(district_name)
  
  lapply(district_vec, function(dist){
    
    formula <- paste0(
      response_var,
      "~",
      paste(independent_var_vec, collapse = "+")
    )
    
    model <- lm(as.formula(formula), data |> filter(district_name == dist))
    
    coef_df <- summary(model)$coefficients |>
      as.data.frame() |>
      tibble::rownames_to_column("var") |>
      rename(
        estimate = Estimate,
        std_error = `Std. Error`,
        pr_t = `Pr(>|t|)`
      ) |>
      filter(var != "(Intercept)")
    
    # SAFE NAME CONVERSION — NEVER RESULTS IN SIZE 0 ERROR
    coef_df <- coef_df |>
      mutate(
        var = sapply(var, function(v){
          out <- variable_long_names |> filter(variable == v) |> pull(long_name)
          if(length(out) == 0) return(v)
          return(out)
        })
      )
    
    coef_df <- coef_df |>
      mutate(
        sig_star = case_when(
          pr_t < 0.01 ~ "***",
          pr_t < 0.05 ~ "**",
          pr_t < 0.10 ~ "*",
          TRUE ~ ""
        ),
        display = paste0(
          round(estimate, 2), " ", sig_star, "<br>(",
          round(std_error, 2), ")"
        ),
        district_name = dist,
        adj_r_sq = summary(model)$adj.r.squared |> round(4),
        rmse = sqrt(mean(model$residuals^2)) |> round(2)
      )
    
    return(coef_df)
  }) |> 
    bind_rows()
}

formula <- paste0("student_teacher_ratio~", paste(variable_vec, collapse = "+"))

lapply(district_vec, function(x){
  
  district <- x
  
  training_data |> 
    filter(district_name == district) -> current_data
  
  current_data <- na.omit(current_data[, c("student_teacher_ratio", variable_vec) ])
  
  model <- lm(as.formula(formula), current_data)
  step(model, direction = "both", trace = 0)
}) |> 
  c() -> district_models

names(district_models) <- district_vec

formula_district_dummy <- paste0(formula, "+format(district_name)")

district_dummy_lm <- lm(as.formula(formula_district_dummy), data = data) 


summary(district_dummy_lm)$coefficients |>
  as.table() |> 
  as.data.frame() |> 
  as.tibble() |>
  pivot_wider(names_from = "Var2", values_from = "Freq") |>
  clean_names() |> 
  filter(var1 != "(Intercept)") |> 
  
  mutate(sig_star = case_when(
    pr_t < 0.005 ~ "***",
    pr_t < 0.05 ~ "**",
    pr_t < 0.1 ~ "*",
    .default = ""
  )) |> 
  
  rowwise() |> 
  mutate(display = paste0(as.character(round(estimate, 3)), " ", sig_star, "<br>(", as.character(round(std_error, 2)), ")")) |> 
  ungroup() |> 
  filter(!grepl("^format\\(district_name\\)", var1)) |> 
  arrange(pr_t) -> district_dummy_regression 


```

**LASSO**

Least Absolute Shrinkage and Selection Operator is a regression method that predicts an outcome and picks out variables that really matter. With a regular OLS regression you end up keeping every variable in model even if many of them are not important or contribute very little. LASSO however adds a penalty. This penalty forces the model to shrink the coefficients of less useful variables, if the penalty is tough enough it can push some coefficients down to zero, kicking out those variables from the model. LASSO is useful when you have many predictors and want a model that is cleaner and easier to understand with only the important variables.

LASSO is useful mainly for:

1.  Variable selection : It identifies which predictors genuinely matter. It gives a model easier to interpret by removing unnecessary variables.
2.  Handle high dimensional data: LASSO works when the number of variables is large or even if they are highly correlated. OLS becomes unstable in such cases but LASSO works here and prevents over fitting.
3.  Prediction accuracy: By keeping only relevant variables the model avoids noise and learns a pattern that generalizes better to new data.

Example: Imagine you want to understand what affects students’ exam scores.\
You collect 50 different variables: study hours, sleep, tuition, attendance, stress levels, screen time, diet quality, parental education, internet speed, and many more.

If you run a normal OLS regression, the model uses all 50 variables even if only 4 or 5 actually matter. This makes the model complicated, noisy, and difficult to interpret.

Now if we add a rule: “Every extra variable you keep in the model costs you a small penalty.” This is what LASSO does, Because of that penalty, the model now prefers to keep only the variables that provide strong predictive power. Variables that contribute very little get shrunk toward zero, and many coefficients become exactly zero. This is gives a much cleaner model.

But how does LASSO know which are the variables that really matter? LASSO identifies important variables by combining prediction accuracy with a penalty that grows with the absolute size of coefficients unlike OLS, which only minimizes prediction error, LASSO minimizes “prediction error + lambda (penalty level) × (sum of absolute coefficients)”, meaning every additional variable carries a cost. If a variable is weak, because it adds noise, overlaps with other predictors, or barely relates to the outcome its contribution to prediction is too small to justify the penalty, so LASSO shrinks its coefficient to zero and effectively removes it. Strong variables survive because shrinking them would noticeably worsen prediction, making it worthwhile to pay the penalty and keep them in the model. Behind the scenes, glmnet tests many penalty levels, from very high ones that keep almost nothing to low ones that resemble OLS, and cross-validation (the n folds) chooses the lambda that predicts best on unseen data.

Lambda (λ) is the penalty strength in LASSO, it controls how harshly the model punishes large coefficients. When λ is small (close to zero), the penalty is weak, and LASSO behaves almost like ordinary OLS, keeping most variables. When λ becomes large, the penalty becomes strong, and LASSO aggressively shrinks many coefficients toward zero, eliminating weaker variables entirely. During cross-validation, glmnet tries many λ values and chooses the one that gives the best prediction accuracy on unseen data.

Alpha (α) controls what type of regularization the model applies. When α = 1, the model becomes pure LASSO, which performs variable selection by shrinking some coefficients all the way to zero. When α = 0, the model becomes Ridge regression, which shrinks coefficients but never drops variables. Values between 0 and 1 give Elastic Net, a mix of both methods. Alpha decides how aggresively the model will drop variables. Since we want to see which variables truly matter at the state level, α = 1 is used so that LASSO can select a smaller, more interpretable set of predictors.

n-fold cross-validation is the method LASSO uses to choose the best penalty strength. The idea is to divide the data into n equal parts, train the model on (n-1) of them, and test it on the remaining part. This process repeats n times so that every part gets used once for testing. The model then picks the lamda value that performs best on average across all folds. Using nfolds prevents overfitting and ensures that the chosen lamda works well on unseen data. I have used nfolds = 10 meaning the model is evaluated 10 times on different splits, which is a standard and reliable choice in machine learning.

```{r packages }
library(glmnet)
library(tidyverse) 
library(gt)

```

```{r Matrix}

y <- data$student_teacher_ratio

# Predictor matrix (remove NAs by filtering rows where y is not NA)
clean_df <- data %>% 
  select(student_teacher_ratio, all_of(variable_vec)) %>% 
  mutate(across(everything(), as.numeric)) %>% 
  drop_na()
y <- clean_df$student_teacher_ratio
x <- clean_df %>% select(-student_teacher_ratio) %>% as.matrix()



```

```{r cross validation}
set.seed(123)

lasso_cv <- cv.glmnet(
  x,
  y,
  alpha = 1,      # LASSO
  nfolds = 10,
  standardize = TRUE
)


```

```{r best model}
lasso_lambda <- lasso_cv$lambda.min     # lambda that gives minimum CV error
lasso_model <- glmnet(x, y, alpha = 1, lambda = lasso_lambda)


```

```{r selected variables,  results="hide"}
coef_df <- coef(lasso_model) %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column("variable")

# Automatically detect the glmnet coefficient column (here it's "s0")
coef_col <- setdiff(colnames(coef_df), "variable")

coef_df <- coef_df %>% 
  rename(coefficient = all_of(coef_col))
coef_df %>% 
  filter(coefficient != 0)

```

The below table shows the variables chosen by LASSO with the alpha 1, nfolds 10 and the lambda value that gives the minimum cross validation error. So here the LASSO penalty parameter - lamda was not chosen manually, instead I used 10 fold cross validation to automatically select the lamda that minimises the prediction error. The value returned by lambda.min was then used to fit the final LASSO model. This is done so that the amount of shrinkage is data driven and improves prediction accuracy.

```{r SELECTED VARIABLES }

# Extract coefficients into a clean data frame
coef_df <- coef(lasso_model) %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column("variable")

# Detect glmnet coefficient column (s0, s1, s2 etc.)
coef_col <- setdiff(colnames(coef_df), "variable")

coef_df <- coef_df %>% 
  rename(coefficient = all_of(coef_col))

# Filter variables selected by LASSO (non-zero coefficients)
lasso_selected <- coef_df %>% 
  filter(coefficient != 0) %>% 
  # Optional: round for readability
  mutate(coefficient = round(coefficient, 4)) %>% 
  # Remove intercept row if you don't want it
  filter(variable != "(Intercept)") 

# Display as a formatted GT table
lasso_selected %>%
  gt() %>%
  tab_header(
    title = md("**Variables Selected by LASSO Regression**"),
    subtitle = "Non-zero coefficients under chosen penalty"
  ) %>%
  
  # Steelblue background for header
  tab_style(
    style = list(
      cell_fill(color = "steelblue"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_title(groups = c("title", "subtitle"))
  ) %>%
  
  cols_label(
    variable = "Variable",
    coefficient = "Coefficient"
  ) %>%
  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_column_labels())
  ) %>%
  
  tab_options(
    table.font.size = 22,
    table.border.bottom.style = "hidden"
  )

```

**COMPARING THE LASSO MODEL WITH THE STATE LEVEL DISTRICT DUMMIES OLS MODEL**

```{r}

formula_district_dummy <- paste0(
  formula,
  " + format(district_name)"
) 
district_dummy_lm <- lm(as.formula(formula_district_dummy), data = data) 


summary(district_dummy_lm)$coefficients |>
  as.table() |> 
  as.data.frame() |> 
  as.tibble() |>
  pivot_wider(names_from = "Var2", values_from = "Freq") |>
  clean_names() |> 
  filter(var1 != "(Intercept)") |> 
  
  mutate(sig_star = case_when(
        pr_t < 0.005 ~ "***",
        pr_t < 0.05 ~ "**",
        pr_t < 0.1 ~ "*",
        .default = ""
      )) |> 
  
  rowwise() |> 
  mutate(display = paste0(as.character(round(estimate, 3)), " ", sig_star, "<br>(", as.character(round(std_error, 2)), ")")) |> 
  ungroup() |> 
  filter(!grepl("^format\\(district_name\\)", var1)) |> 
  arrange(pr_t) -> district_dummy_regression
 


ols_coef <- summary(district_dummy_lm)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  filter(!grepl("^format\\(district_name\\)", variable)) %>%   
  filter(variable != "(Intercept)") %>%
  rename(
    estimate = Estimate,
    p_value = `Pr(>|t|)`
  )

```

```{r}
lasso_coef <- coef(lasso_model) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  rename(coefficient = s0) %>%    # your column name was s0
  filter(variable != "(Intercept)")
lasso_selected <- lasso_coef %>%
  filter(coefficient != 0)

```

```{r}

# OLS results from district-dummy regression
ols_df <- district_dummy_regression %>%
  select(variable = var1, estimate)

# LASSO results extracted earlier
coef_df <- coef(lasso_model) %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  rownames_to_column("variable")

coef_col <- setdiff(colnames(coef_df), "variable")

coef_df <- coef_df %>%
  rename(coefficient = all_of(coef_col))

# Remove intercept
coef_df <- coef_df %>% filter(variable != "(Intercept)")

# Merge OLS + LASSO
comparison_df <- ols_df %>% 
  left_join(coef_df, by = "variable") %>%
  mutate(
    selected_by_ols = !is.na(estimate),
    selected_by_lasso = coefficient != 0
  )



comparison_df %>%
  mutate(
    estimate = round(estimate, 4),
    coefficient = round(coefficient, 4),
    selected_by_ols = ifelse(selected_by_ols, "Yes", "No"),
    selected_by_lasso = ifelse(selected_by_lasso, "Yes", "No")
  ) %>%
  gt() %>%
  tab_header(
    title = md("**Comparison of State-level OLS vs. LASSO**"),
    subtitle = "Which variables each method selects"
  ) %>%
  
  # Steelblue header styling
  tab_style(
    style = list(
      cell_fill(color = "steelblue"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_title(groups = c("title", "subtitle"))
  ) %>%
  
  cols_label(
    variable = "Variable",
    estimate = "OLS Coefficient",
    coefficient = "LASSO Coefficient",
    selected_by_ols = "Selected by OLS?",
    selected_by_lasso = "Selected by LASSO?"
  ) %>%
  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_column_labels())
  ) %>%
  
  tab_options(
    table.font.size = 22,
    table.border.bottom.style = "hidden"
  )


```

The comparison table shows that the state level OLS and the LASSO model gave the same variables as the output. Even after applying an automatic variable selection method with a penalty LASSO did not shrink any coefficient to zero. This could suggest that OLS and LASSO agree on the direction and relevance of all variables. Every variable has a non zero coefficient, a non zero LASSO coefficient. This indicated that the predictors have a consistent, stable relationship with student teacher ration at the state level. No variable appears to be purely noise under the chosen penalty parameter. It could be that the LASSO penalty was not strong enough to force sparsity. However, we can see some heavier shrinkage in some variables than others, based on their predictive capacities and how correlated they are with other predictors. Public information board for instance moves coefficient from -0.0632 in OLS to -0.9491 in LASSO. Overall, this suggests that the determinants of student teacher ratio in primary school are multi-dimensional, spanning across middle school access, health access, infrastructure, connectivity, socio economic services, public services etc.

**CHANGING THE PENALTY PARAMETER (LAMBDA)**

```{r}
### Choose new alpha and lambda here:
alpha_new <- 1      
lambda_new <- 0.5  



### Fit model with new settings
lasso_model_new <- glmnet(
  x,
  y,
  alpha = alpha_new,
  lambda = lambda_new,
  standardize = TRUE
)

### Extract coefficients
new_coef_df <- coef(lasso_model_new) %>%
  as.matrix() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable")

new_coef_col <- setdiff(colnames(new_coef_df), "variable")

new_coef_df <- new_coef_df %>%
  rename(coefficient = all_of(new_coef_col)) %>%
  filter(variable != "(Intercept)")

### Merge with OLS
comparison_new <- ols_df %>%
  left_join(new_coef_df, by = "variable") %>%
  mutate(
    selected_by_ols = !is.na(estimate),
    selected_by_lasso = coefficient != 0
  )

### Clean + display table
comparison_new %>%
  mutate(
    estimate = round(estimate, 4),
    coefficient = round(coefficient, 4),
    selected_by_ols = ifelse(selected_by_ols, "Yes", "No"),
    selected_by_lasso = ifelse(selected_by_lasso, "Yes", "No")
  ) %>%
  gt() %>%
  tab_header(
    title = md(paste0("**Comparison: OLS vs. LASSO (α = ", alpha_new,
                      ", λ = ", lambda_new, ")**")),
    subtitle = "Effect of tuning parameters on variable selection"
  ) %>%
  
  # Steelblue header styling
  tab_style(
    style = list(
      cell_fill(color = "steelblue"),
      cell_text(color = "white", weight = "bold")
    ),
    locations = cells_title(groups = c("title", "subtitle"))
  ) %>%
  
  cols_label(
    variable = "Variable",
    estimate = "OLS Coefficient",
    coefficient = "LASSO Coefficient",
    selected_by_ols = "Selected by OLS?",
    selected_by_lasso = "Selected by LASSO?"
  ) %>%
  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_column_labels())
  ) %>%
  
  tab_options(
    table.font.size = 22,
    table.border.bottom.style = "hidden"
  )


```

When the penalty parameter - lamda is increased to 0.5, LASSO becomes much more aggressive in shrinking coefficients towards zero. Unlike the earlier cross validated model where LASSO kept all the variables the higher penalty forces the model to become sparse, selecting only a subset of the predictors. The comparison table shows a dramatic difference between OLS and LASSO. With lamda is 0.5 the model becomes way more sensistive and eliminates variables that OLS considered significant. The set of variables that survived the penalty are:

-   `availability_of_middle_school`

-   `health_centre`

<!-- -->

-   `share_grd_pg`

<!-- -->

-   `availability_of_govt_degree_coll`

<!-- -->

-   `ln_labour_budget_per_capita`

<!-- -->

-   `share_insurance_receiving`

<!-- -->

-   `share_farming_activity`

<!-- -->

-   `is_bank_available`

<!-- -->

-   `share_hhd_ssh`

These are the variables whose association with student–teacher ratio is strong enough that even a bigger penalty cannot shrink their coefficients to zero. These predictors likely carry the most consistent, independent predictive information. Even among the variables that survived selection, their coefficients become much smaller in magnitude compared to OLS. The shrinkage reflects LASSO's general bias towards models that avoids overfiting.

Many variables are eliminated not because they have no effect, but because:

-   They are highly correlated with other predictors. Their effect is not strong enough to justify their inclusion under a large penalty. For example:

-   `is_primary_school_with_toilet`

-   `is_village_connected_to_all_weat`

-   `is_primary_school_with_electrici`

These infrastructure variables may be strongly correlated with each other, so LASSO keeps only the dominant ones and drops the rest.

This comparison clearly shows:

-   OLS keeps almost everything because it has no penalty

-   LASSO with cross-validated λ keeps everything but shrinks them slightly

-   LASSO with λ = 0.5 (strong penalty) forces a much simpler model

Increasing lambda to 0.5 dramatically changed the variable selection. Under this stronger penalty, LASSO removed almost half of the predictors that OLS considered statistically significant. This happened because a higher lambda forces the model to keep only the strongest, most stable predictors, while shrinking correlated or weaker variables to zero. As a result, the lambda = 0.5 LASSO model is far sparser and more interpretable than both OLS and the cross-validated LASSO model. However, it may also lose information, as several dropped variables still had meaningful coefficients in OLS. This illustrates the core tradeoff in LASSO: simpler models versus possible loss of explanatory detail.

The LASSO model differs substantially from the state-level OLS regression with district dummies that we estimated in the group project. In the OLS model, every variable entered the regression and almost all predictors appeared statistically significant, largely because OLS does not penalize model complexity. In contrast, LASSO introduces a penalty that forces the model to keep only the strongest and most stable predictors. Under the cross-validated lambda, LASSO retained all variables but shrank their coefficients noticeably, showing that many predictors carried weaker signals than OLS suggested. When the penalty was increased (lambda = 0.5), the contrast became even more striking: LASSO eliminated almost half the variables that OLS considered significant, setting their coefficients exactly to zero. Variables such as connectivity, broadband availability, school toilet facilities, piped water, electricity in primary schools, immunization and newborn health indicators, and several schooling infrastructure variables were removed entirely because their predictive contribution was either weak or redundant once other correlated variables were included. The variables that survived the higher penalty, such as availability of middle schools and degree colleges, health centres, share of graduates and postgraduates, farming activity, SHG participation, and insurance coverage, are those with the strongest independent association with student–teacher ratio. These differences show that while OLS highlights statistical significance, LASSO focuses on predictive strength, revealing which variables genuinely matter once the model is penalized for complexity.
